---
title: "11 Multiple Regression Analysis with Qualitative Regressors"
author: "Mohammadamin Firouzi"
date: "2025-04-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Multiple Regression Analysis with Qualitative Regressors

If a single dummy variable is used in a model, its coefficient represents the difference in the intercept between groups.

```{r message=FALSE, warning=FALSE}

library(dplyr)

data(wage1, package = 'wooldridge')
wage_vs_female = lm(wage ~ female, wage1)
summary(wage_vs_female)

mean = wage1 |>
  group_by(female) |>
  summarize(wage_mean = mean(wage))

mean
```

```{r message=FALSE, warning=FALSE}

wage_vs_educ_female = lm(wage ~ educ + female, wage1)
summary(wage_vs_educ_female)
```

Its estimated coefficient of −2.27 indicates that on average, a woman makes \$2.27 per hour less than a man with the same education.

```{r message=FALSE, warning=FALSE}

lwagereg = lm(log(wage) ~ married*female + educ + exper + I(exper^2) + tenure + I(tenure^2), wage1)
summary(lwagereg)
```

Relative to the reference group of single males with the same education, experience, and tenure, married males make about 21.3% more (the coefficient of married), and single females make about 11.0% less (the coefficient of female). The coefficient of the interaction term implies that married females make around 30.1%-21.3%=8.7% less than single females, 30.1%+11.0%=41.1% less than married males, and 30.1%+11.0%-21.3%=19.8% less than single males.

#### Logical variables:

```{r message=FALSE, warning=FALSE}

wage1$female = as.logical(wage1$female)
table(wage1$female)

summary(lm(wage ~ female + educ, data = wage1))

wage1$female = as.numeric(wage1$female)
```

#### Factor variables:

```{r message=FALSE, warning=FALSE}

library('AER')
data(CPS1985)

head(CPS1985)

attach(CPS1985)

table(gender)
table(occupation)

regression = lm(wage ~ education + experience + gender + occupation)
summary(regression)
```

#### Redefine the reference category:

```{r message=FALSE, warning=FALSE}

occupation = relevel(occupation, 'office')

regression2 = lm(wage ~ education + experience + gender + occupation)
summary(regression2)
```

#### Breaking a Numeric Variable Into Categories:

```{r message=FALSE, warning=FALSE}

data(lawsch85, package = 'wooldridge')
View(lawsch85)

cutpts = c(0, 10, 25, 40, 60, 100, 175)

lawsch85$rankcat = cut(lawsch85$rank, cutpts)

table(lawsch85$rankcat)

lawsch85$rankcat = relevel(lawsch85$rankcat, '(100,175]')

regression3 = lm(log(salary) ~ rankcat + LSAT + GPA + log(libvol) + log(cost), data = lawsch85)
summary(regression3)

(exp(regression3$coefficients[2]) - 1)
```
