---
title: "1 Basics"
author: "Mohammadamin Firouzi"
date: "2025-02-16"
output: html_document
---

# **Basics**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### R as a calculator

```{r, warning=FALSE, message = FALSE}

sqrt(4)
10 + 7.5
10 * (1.5 - 0.25)^2
```

------------------------------------------------------------------------

#### Important R function

```{r, warning=FALSE, message = FALSE}

abs(-10) # absolute value

round(142.21, 1) # round 142.21 to 1 digit
round(24.23152, 4) # round 24.23152 to 4 digits

log(6) # ln(6)
log(6, 10) # logarithm to base 10

exp(123) # exponential function e^123

factorial(6) # factorial 6!
factorial(10) # factorial 10!

choose(10, 5) # binomial coefficient
choose(123, 100)

#n()  returns the size of a group
#sum(!is.na())  count the number of non-missing values
#n_distinct()  count the number of unique values
```

------------------------------------------------------------------------

#### Objects

We will often want to store results of calculations to reuse them later. For this, we can work with basic objects.

R is case sensitive, so x and X are different object names.

```{r, warning=FALSE, message = FALSE}

x = 6
# or
x <- 6
x

(z = 1) # Generate object with immediate display.

log(x)
x^2
factorial(x)
```

------------------------------------------------------------------------

```{r, warning=FALSE, message = FALSE}

ls() # A list of all currently defined object names can be obtained using ls().
rm(x) # Removing x or any other object you want.
rm(list = ls()) # Removing all objects.
```

------------------------------------------------------------------------

#### Vectors

To define a vector, we can collect different values using c(value1,value2,...).

```{r, warning=FALSE, message = FALSE}

y = c(2, 4, 6, 8, 10) 
print(y)

(z = c(1, 3, 5, 7, 9))

sqrt(y)
abs(y)
log(y)

(yz = y * z)
```

#### R functions specifically for vectors

```{r, warning=FALSE, message = FALSE}

max(yz) # Largest value in vector 
min(yz) # Smallest value in vector

a = c(5, 1, 3, 9, 7) 
sort(a) # Sort the elements of vector

sum(yz) # Sum of the elements of vector 
prod(yz) # Product of the elements of vector 
length(yz) # Number of elements in vector

numeric(5) # Vector with 5 zeros 
rep(1, 5) # Vector with 5 equal elements 1 
rep(10, 10)

seq(10) # Sequence from 1 to 10 
seq(100) # Sequence from 1 to 100 
# or
1:10
1:100

seq(5, 10) # Sequence from 5 to 10 
seq(65, 100) # Sequence from 65 to 100 
#or 
5:10 
65:100

seq(5, 17, 4) # Sequence from 5 to 15 in steps 3.
seq(10, 100, 10) # Sequence from 10 to 100 in steps 10.

names = c("Atousa", "Mani", "Parsa") 
names

```

------------------------------------------------------------------------

#### Logical operation

```{r, warning=FALSE, message = FALSE}

x = 1
y = 2

x == y  # x is equal to y 
x < y # x is less than y 
x <= y # x is less than or equal to y
x > y # x is greater than y
x >= y # x is greater than or equal to y
x != y # x is NOT equal to y

!x # NOT x (i.e. x is FALSE)
x|y # Either x or y is TRUE (or both)
x&y # Both x and y are TRUE
```

------------------------------------------------------------------------

#### Factors

Many economic variables of interest have a qualitative rather than quantitative interpretation. They only take a finite set of values and the outcomes don’t necessarily have a numerical meaning. Instead, they represent qualitative information.

```{r, warning=FALSE, message = FALSE}

gender = c(0, 0, 1, 1, 0, 0, 0, 0, 0, 1) # 0 = "male", 1 = "female"
gender_factor = factor(gender, labels = c("male", "female"))

gender
gender_factor
```

------------------------------------------------------------------------

#### Naming

```{r, warning=FALSE, message = FALSE}

numbers = c(1, 2, 3, 4)
students = c("Atousa", "Mani", "Parsa", "Sadra")
names(numbers) = students

numbers
```

#### Indexing

Indices by number:

```{r, warning=FALSE, message = FALSE}

numbers[2]
numbers[2:4]
```

Indices by name:

```{r, warning=FALSE, message = FALSE}

numbers['Atousa']
```

Logical indices:

```{r, warning=FALSE, message = FALSE}

numbers[ numbers >= 3]
```

------------------------------------------------------------------------

#### Matrices

Three different ways to define a matrix object from scratch:

```{r, warning=FALSE, message = FALSE}

vector = c(1, 2, 3, 4, 5, 6)
matrix(vector, ncol = 3)
matrix(vector, ncol = 2)

r1 = c(1, 2, 3)
r2 = c(7, 8, 9)
r3 = c(4, 5, 6)

rbind(r1, r2, r3)
cbind(r1, r2, r3)
```

Diagonal and identity matrices:

```{r, warning=FALSE, message = FALSE}

diag( c(7, 8, 10) )
diag(10)

N = cbind(r1, r2, r3)
N
diag(N)
```

Indexing:

```{r, warning=FALSE, message = FALSE}

X = rbind(r1, r2, r3)
X

X[3, 3]
X[2,]
X[,1]
X[,c(1, 3)]
X[c(1, 2), c(1, 3)]
```

Matrix operators

```{r, warning=FALSE, message = FALSE}

X
N
X * N

X %*% N # Matrix multiplication

t(X) #Transpose
t(N) 

A = matrix(c(1, -10, 12, 100, 32, -7, 56, 0, -8), nrow = 3)
A
solve(A) # Inverse
```

------------------------------------------------------------------------

#### List

In R, a list is a generic collection of objects. Unlike vectors, the components can have different types. Each component can (and in the cases relevant for us will) be named.

```{r, warning=FALSE, message = FALSE}

listone = list(name = 'A', name1 = 'B')
listone

listtwo = list(name = 'A', s = seq(1, 10, 2), mat = diag(2))
listtwo

names(listtwo) # Vector of names

listtwo$name
listtwo$mat
```
